---
title: "Introduction to dplyr -  make data manipulation easier"
output: html_document
---

## Description of dplyr
* created by Hadley Wickham (Chief Scientist of Rstudio, author of `ggplot`, `tidyr`, `broom`)
* Great for data exploration and transformation
* Intuitive to write and easy to read, especially when using the "chaining (%>%)" syntax 
* Faster than basic R approach

## dplyr functionality
* Five basic verbs: `filter`, `select`, `arrange`, `mutate`, `summarise` (plus `group_by`)
* Lots of handy functions (i.e., `mutate_each`, `summarise_each`, `n_distinct`)
* Joining (merging) data frames: `left_join`, `right_join`, `inner_join`, `full_join`, `semi_join`, `anti_join`
* [Window function](http://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html) for calculating ranking(like `min_rank`), offsets(`lead` and `lag`), and more (not covered below; see [vignette for window-functions](https://github.com/hadley/dplyr/blob/master/vignettes/window-functions.Rmd))
* Connecting to Databases using SQLite, PostgreSQL/Redshift, MySQL/MariaDB, BigQuery, MonetDB (not covered below; see [vignette for database](http://rpubs.com/justmarkham/dplyr-tutorial))

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                     warning=FALSE, message=FALSE)
```

* load packages and an example dataset
```{r}
library(dplyr)
```

```{r include=FALSE}
setwd("C:/Users/S00174624/Google Drive/IPPE_Rcourse_Jiesi")
pisa<-read.csv("PISA12-AUS-motivation.csv", stringsAsFactors = FALSE)
```

```{r eval=FALSE}
pisa<-read.csv("PISA12-AUS-motivation.csv", stringsAsFactors = FALSE)
```

* `tbl_df` creates a "local data frame"
* Local data frame is simply a wrapper for a data frame that is easier to examine.

```{r}
# convert to local data frame
pisa <- tbl_df(pisa)
# printing only shows 10 rows and as many columns as can fit on your screen
pisa
```

### Viewing printing more output

```{r eval=FALSE}
# you can specify that you want to see more rows
print(pisa, n=20)
# slice() certain row numbers
slice(pisa, 1:5)
# specify that you want to see ALL rows (don't run this!)
print(pisa, n = Inf)
# specify that you want to see ALL columns
print(pisa, width = Inf)
# convert to a normal data frame to see all of the columns
#data.frame(head(pisa))

# set option to set numbers of columns and rows you want to see
options(dplyr.width = 20, dplyr.print_min = 20)
# reset options (or just close R)
options(dplyr.width = NULL, dplyr.print_min = 10)
```

## filter: Keep rows matching criteria

* dplyr approach is simpler to write and read
* Command structure (for all dplyr verbs):
    * first argument is a data frame
    * return value is a data frame
    * nothing is modified in place

```{r eval=FALSE}
# base R approach to view all students in catholic school in NSW
pisa[pisa$Schtype==1 & pisa$State=="NSW", ]# Base R approach forces you to repeat the data frame's name
# schtype: 1. catholic, 2. government, 3. independent
```

```{r}
# dplyr approach
filter(pisa, Schtype==1, State=="NSW") # using comma or ampersand to represent AND condition 
```

```{r eval=FALSE}
# to view all students in NSW and VIC 
filter (pisa, State=="NSW" | State=="VIC") # using pipe to present OR condition
# you can also use %in% operator
filter (pisa, State %in% c("NSW", "VIC"))
```
* other examples for selecting columns 
```{r eval=FALSE}
# between() is a concise alternative for determing if numeric values fall in a range
# to view students having math scores from 300 to 500
filter(pisa, between(PV1MATH, 300, 500))

# is.na() can also be useful when filtering
# remove missing value from ST42Q02
filter(pisa, !is.na(ST42Q02))
```

## select: Pick columns by name

```{r eval=FALSE}
# base R approach to select school sype, state and math achievement
pisa[, c("Schtype", "State", "PV1MATH")]
```

```{r}
# dplyr approach
select(pisa, Schtype, State, PV1MATH) 
```
### other examples for selecting columes by name
```{r eval=FALSE}
select(pisa, Schtype:PV1MATH)      # From Schtype through PV1MATH.
select(pisa, -State, -PV1MATH)     # hide those two variables.
select(pisa, contains("MAT"))      # Gets variables containing "MAT"
select(pisa, starts_with("PV"))    # Gets all vars staring with "PV".
select(pisa, ends_with("MAT"))     # All vars ending with "MAT".
select(pisa, matches("^ST\\d{2}"))  # Matches any regular expression.
```

## "Chaining" or "Pipelining" - the key feature of dplyr

* Can write commands in a natural order by using the `%>%` infix/pipe operator imported from [magrittr](https://github.com/smbache/magrittr) package (which can be pronounced as "then")
```{r eval=FALSE}
# take the thing on the left hand-side and insert it as the first argument on the right hand-side
# x %>% f(y) equal to f(x,y)
# x %>% f(y) %>% g(z) equal to g(f(x,y),z)
```

```{r results='hide'}
# nesting method to select State and PV1MATH columns and filter for PV1MATH over 600
filter(select(pisa, State, PV1MATH), PV1MATH > 600)
```

```{r}
# chaining is an alternative way to write your R code
pisa %>%
    select(State, PV1MATH) %>%
    filter(PV1MATH > 600)
```    

## rename: specific columns
```{r results='hide'}
# Reman ESCS to SES
rename(pisa, SES = ESCS)
```

## arrange: Reorder rows

```{r results='hide'}
# base R approach to select State and PV1MATH columns and sort by PV1MATH
pisa[order(pisa$PV1MATH), c("State", "PV1MATH")]
```

```{r}
# dplyr approach
pisa %>%
    select(State, PV1MATH) %>%
    arrange(PV1MATH) #use `desc` for descending: arrange(desc(PV1MATH))
```
## mutate: Add new variables

* Create new variables that are functions of existing variables

```{r eval=FALSE}
# base R approach to create a new variable for differece between students' math achivement and science achievement
pisa$diff_ach <- pisa$PV1MATH - pisa$PV1SCIE
pisa[, c("PV1MATH", "PV1SCIE", "diff_ach")]
```

```{r}
# dplyr approach (prints the new variable but does not store it)
pisa %>%
    select(PV1MATH, PV1SCIE) %>%
    mutate(diff_ach = PV1MATH - PV1SCIE)

# store the new variable
pisa <- pisa %>% mutate(diff_ach = PV1MATH - PV1SCIE) 
# `transmute` only keeps the new variables
pisa %>% transmute(diff_ach = PV1MATH - PV1SCIE) 
```

## summarise: Reduce variables to values

```{r}
# summary the mean of math achievement and math self-concept
pisa %>% summarise(mean_math = mean(PV1MATH), mean_msc = mean(SCMAT, na.rm = TRUE))
```
## Combine `mutate` and `summarise` with `group_by`
*  `mutate` and `summarise` are primarily useful with data that has been grouped by one or more variables
``` {r}
# local data frame is the same as the original one except for grouping State
pisa %>% group_by(State) 
```

``` {r}
# view the mean of math achievement and self-concept for each state
pisa %>% 
    group_by(State) %>%
    summarise(mean_math = mean(PV1MATH), mean_msc = mean(SCMAT, na.rm = TRUE))
```

```{r}
#  view the number of total students and number of distinct schools, for each state based on each school type
 pisa %>% 
    group_by(State,Schtype) %>%
    summarise(stu_count = n(), school_count = n_distinct(SCHOOLID)) 
# function `n()` counts the number of rows in a group
# function `n_distinct(vector)` counts the number of unique items in that vector
 	
```
* `summarise_each` allows you to apply the same summary function to multiple columns at once
```{r}
# view  max and min achievement scores of three school subjects (reading, math, science) for each state based on eath school type
pisa %>%
 	group_by(State,Schtype) %>%
 	summarise_each(funs(max, min), PV1MATH:PV1SCIE) %>%
  print (width = Inf)
# if varibles with missing value, summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), PV1MATH:PV1SCIE))
``` 	
* `mutate_each` allows you to apply the same mutate function to multiple columns at once to create new variables

```{r }
#for easy presentation, I reduce the size of data set in this case
pisa1 <- pisa %>% select (SCHOOLID,SCMAT,PV1MATH:PV1SCIE) 
# create average school math and science achievement for each student
pisa1 %>%
 	group_by(SCHOOLID) %>% 
 	mutate_each(funs(mean_ach = mean), -SCMAT, -PV1READ)
#Here a small bug: when only 1 function is applied, dplyr will keep the orignal names (replace the existing variales)
```

```{r}
# What you can do to avoid this is specifying names of each variable manually 
pisa1 %>%
	group_by(SCHOOLID) %>% 
	mutate_each(funs(mean), mean_math = PV1MATH, mean_scie = PV1SCIE, -SCMAT, -PV1READ) 
```

```{r}
# when 2 functions are applied, variable names works fine
# create the standardized and centered scores of each student's reading, math and science achievement 
pisa1 %>%
    mutate_each(funs(std = scale(.), cent = scale(., scale = FALSE)), -SCMAT, -SCHOOLID) %>%
    print(width = Inf)
```
## `do` allows you to apply one unnamed argument 
```{r}
# for each of school type, do simple regression of math achievement on math self-concept 
pisa %>% 
  	group_by(Schtype) %>%
  	do(
   	output = lm(PV1MATH ~ SCMAT, data = .)
 			) 
# store to a new local data frame
models <- pisa %>% 
  			group_by(Schtype) %>%
  			do(
   			 output = lm(PV1MATH ~ SCMAT, data = .)
          )
```

```{r}
# create a function to extract path coefficients, standard error, t-value and p-value
coef_df <- function(x) {
  temp <- coef(summary(x))
  colnames(temp) <- c("est", "se", "t", "P")
  data.frame(coef = rownames(temp), temp)
}
models %>% do(coef_df(.$output))
```

```{r}
# the easy way to extract similar output by using `tidy` function in `broom` package
#install.packages("broom") 
library(broom)
models %>% do(tidy(.$output)) #`tidy` function can be used for different types of regression models
```

## Joining (merging) data frames

```{r}
# create two simple data frames
a <- data.frame(color = c("black","blue","red"), number = 4:6, stringsAsFactors = FALSE)
b <- data.frame(color = c("black","blue","pink"), size = c("S","M","L"), stringsAsFactors = FALSE)

# the three most common joins
# only include observations found in both "a" and "b" (automatically joins on variables that appear in both tables)
inner_join(a, b)
```

```{r}
# include observations found in either "a" or "b" 
full_join(a, b)
```

```{r}
#include all observations found in "a"
left_join(a, b)
```
* see [vignette for two-tables](https://github.com/hadley/dplyr/blob/master/vignettes/two-table.Rmd) for `right_join`, `left_join`, `semi_join` and `anti_join`

## Combining data frames
```{r}
# create a simple dataframe
mydata <- data.frame(
  x = 1:3,
  y = c("a", "b", "c"),
  stringsAsFactors = FALSE) 
```
```{r}
# when combinding column-wise,the number of rows must match, but row names are ignored
bind_cols(mydata, data.frame(z = 3:1)) 
# when combinding row-wise, both the number and names of columns much match
bind_rows(mydata, data.frame(x = 10, y = "z")) 
```
* use`rbind_all` to combind multiple data frames like `do.call(rbind, )`

## References and Resourse
* [dplyr reference manual and vignettes](http://cran.r-project.org/web/packages/dplyr/)
* [dplyr tutorial by Hadley Wickham](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a) at the [useR! 2014 conference](http://user2014.stat.ucla.edu/): excellent, in-depth tutorial with lots of example code (Dropbox link includes slides, code files, and data files)
* [dplyr tutorial One by Kevin Markham](https://www.youtube.com/watch?v=jWjqLW-u3hc&list=PL5-da3qGB5IBM6DdZ-ZV8mA0oV0U56xaq)
* [dplyr tutorial Two by Kevin Markham](https://www.youtube.com/watch?v=2mh1PqfsXVI&list=PL5-da3qGB5IBM6DdZ-ZV8mA0oV0U56xaq)
* [dplyr Cheat Sheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
* [dplyr GitHub repo](https://github.com/hadley/dplyr) 
